---
title: "Data cleaning"
date: "2017-10-31"
editor_options: 
  chunk_output_type: console
---

```{r load-libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(readxl)
library(pander)
library(stringr)
library(DT)
library(here)
```

## Load data

There are two different versions of the data—an SPSS file and an Excel file—and neither are perfectly identical. These columns have issues:

- `ID`: This columnn is full of NAs in SPSS; it exists in Excel.
- `ACT`, `BIRTH`, `SAT`, `UGA_ID`: These columns are almost full of NAs in SPSS; they kind of exist in Excel—when values do exist, they're separated by commas, like `2,9`, `1,3`, `1, 8, 9, 0`, etc.
- `BIRTH`: This column is almost full of NAs in SPSS; it kind of exists in Excel—when values do exist, they're also separated by commas.
- Most other columns are fine, though some cells in Excel (like row 10 in `EDU_DAD`) have comma-separated values; in SPSS, the last value is used.

```{r load-data}
psm_raw_spss <- read_spss(file.path(here(), "data", "data_raw", "Student PSM Survey Data.sav")) %>% 
  zap_formats() %>% zap_widths()

psm_raw_excel <- read_excel(file.path(here(), "data", "data_raw", "Student PSM Survey Master Data.xlsx"),
                            sheet = "Master Data")
```

#### Data from SPSS

```{r show-spss, results="asis"}
psm_raw_spss %>%
  select(1:5) %>% head() %>% 
  pandoc.table()
```

#### Data from Excel

```{r show-excel, results="asis"}
psm_raw_excel %>%
  select(1:5) %>% head() %>%
  pandoc.table()
```

Because of this weird discrepancy, I use all columnns from the SPSS data (since it takes care of the strange comma-separated values), but I add the ID column from Excel for identification and grouping purposes later. For now, I ignore `ACT`, `BIRTH`, and other columns that behave strangely.

#### Final raw data

```{r final-raw-data, results="asis"}
psm_raw <- bind_cols(
  select(psm_raw_excel, 1),
  select(psm_raw_spss, -1)
)

psm_raw %>% 
  select(1:5) %>% head() %>%
  pandoc.table(missing = "")
```


## Calculate index totals

### Map questions to indexes

There is a simplified codebook that matches each of the questions in the survey to specific items in the four main PSM surveys:

```{r loadl-clean-index, results="asis"}
psm_codebook_simple <- read_excel(file.path(here(), "data", "data_raw", "psm measures_1.xls"),
                            sheet = "Sheet1", col_names = c("variable", "details", "question")) %>%
  mutate(index = ifelse(str_detect(variable, "psm_"), NA, variable)) %>%
  fill(index) %>%
  filter(!is.na(details)) %>%
  mutate(index = recode(index, `Int'l Measure` = "International"),
         variable = str_to_upper(variable),
         variable = recode(variable, PSM_25 = "PSM_25_R", PSM_41 = "PSM_41_R"))

psm_codebook_simple %>% datatable()
```

### Calculate totals

Using this cleaner data, I extract the questions that apply to each of the four indexes:

```{r add-index-totals, results="asis"}
grant_vars <- psm_codebook_simple %>% filter(index == "Grant") %>% pull(variable)
intl_vars <- psm_codebook_simple %>% filter(index == "International") %>% pull(variable)
msp_vars <- psm_codebook_simple %>% filter(index == "MSPB5") %>% pull(variable)
perry_vars <- psm_codebook_simple %>% filter(index == "Perry") %>% pull(variable)

index_vars <- tribble(
  ~Index, ~Columns,
  "Perry", perry_vars,
  "MSPB5", msp_vars,
  "Grant", grant_vars,
  "International", intl_vars
) %>%
  mutate(Variables = Columns %>% map_int(~ length(.)),
         Columns = Columns %>% map_chr(~ paste0("`", ., "`", collapse = ", ")))

index_vars %>% pandoc.table(justify = "lll")
```

I add each of those columns up to create the index for each individual (and calculate the z-score):

```{r calculate-index-totals, results="asis"}
psm_indexes <- psm_raw %>%
  mutate(index_perry = rowSums(select(., one_of(perry_vars))),
         index_msp = rowSums(select(., one_of(msp_vars))),
         index_grant = rowSums(select(., one_of(grant_vars))),
         index_intl = rowSums(select(., one_of(intl_vars)))) %>%
  mutate_at(vars(starts_with("index")), funs(z = scale(.)))

psm_indexes %>% 
  select(ID, starts_with("index")) %>%
  head() %>%
  pandoc.table(split.table = Inf, missing = "")
```

## Save data

Finally, I save a clean version of the data for use elsewhere.

```{r save-clean-data}
write_csv(psm_indexes, file.path(here(), "data", "data_clean", "psm_clean.csv"))
```
